---
title: "Artik"
editor: visual
execute:
  echo: false
---

```{r message=FALSE}
library(tidyverse)
artik <- read_csv("artik.csv")
```
Artik is a town in the Artik Municipality of the Shirak Province of Armenia. 

Now we will explore the fire causes probability in Artik, going through data provided by Tumo. 
The graph below represents the temperature and humidity flow during May-July 2024. As you can see, there is no information in the start of June, because the device was broken.
The danger humidity level was registered from the middle of June to the start of July.

```{r message=FALSE}
library(ggplot2)
library(dplyr)
library(patchwork)
p1 <- artik |> drop_na() |> ggplot(aes(x=time, y=temperature)) + geom_line(color="darkblue") + geom_smooth(color="white") + theme_minimal() + labs(x=NULL, y="Temperatue", title="Temperature and Humidity in Artik")

p2 <- artik |> drop_na() |> ggplot(aes(x=time, y=humidity)) + geom_line(aes(color=if_else(humidity<30,"too_low", "normal")))+ scale_color_manual(values = c("too_low"="red", normal="orange"), guide="none") + theme_minimal() + labs(x="Time", y="Humidity")

combined_plot <- p1/p2

print(combined_plot)
```

Humidity level is considered to be low when being lower than 30 percents. The graph below represents the humidity level when the temperature is above 25 C (May-July 2024). 

```{r message=FALSE}
artik |> drop_na() |> mutate(humidity_level=humidity>30, .before=1) |> filter(temperature>25) |> ggplot(aes(x=time, y=temperature)) + geom_point(aes(color=if_else(humidity<30, "too_low", "normal"))) + scale_color_manual(values=c("too_low"="firebrick1", "normal"="steelblue1"), labels=c("too_low"="Too low", normal="Normal")) + labs(color="Humidity level", x="Time", y="Temperature") + theme_minimal() 
```

### Predicting the spread of fire
We never know when and where will start the fire, so predicting the fire spreading speed and direction would be helpful.
In the graph below we can see the wind speed during May-July 2024. As you can see, there is not danger. The wind speed specifically grows at the end of May and at the start of July.

```{r message=FALSE}
artik |> drop_na() |> ggplot(aes(x=time, y=speed)) + geom_jitter(aes(color=if_else(speed<15,"normal", "too_high"), alpha=0.5))+ scale_color_manual(values = c("normal"="black", too_high="darkred"), labels=c("normal"="Normal Speed", "too_high"="High Speed")) + scale_alpha_continuous(guide = "none") + geom_smooth(color="white") + theme_minimal() + labs(title="Wind speed in Artik", x ="Time", y="Speed", color="Wind speed danger level")
```

```{r message=FALSE}
artik |> drop_na() |> ggplot(aes(x=direction)) + geom_bar(fill="skyblue4") + theme_minimal() + labs(x="Direction", y="Count")
```

